Time and space specifications for stencils

### 2016-03-10 paper.tex
working on description of equality and inequality
work on the multi-set model (missing temporal specs); and start of soundness proof

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}

@@

@@

\newtheorem{theorem}[block]{Theorem}
\newtheorem{lemma}[block]{Lemma}
\newtheorem{proposition}[block]{Proposition}
%\newtheorem{definition}[block]{Definition}

\theoremstyle{definition}
\newtheorem{remark}[block]{Remark}
\newtheorem{example}[block]{Example}
\newtheorem{definition}[block]{Definition}

@@

This paper makes the following contributions:
%
\begin{itemize}
\item introduces a specification language for
stencil computations that captures many common forms
of stencil (Section~\ref{sec:lang});
\item we detail specification inference and checking
algorithsm (Section~\ref{sec:analysis});
\item we evaluate our implementation of the approach
in the CamFort tool for Fortran verification, studying
a number of example programs to assess the usefulness
of this approach.
\dnote{insert results here}
\end{itemize}
%

@@

\begin{figure}
\begin{align*}
spec ::= & \; S  \; \texttt{,} \; spec \, \mid \, S \\[0.5em]
%
S ::= &  \; \texttt{forward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{backward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{centered} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{unspecified} \; [\texttt{dims=}\textit{dims}] \\
& \; \mid \texttt{read-once} \; S \\[0.5em]
\textit{depth} ::= & \; \mathbb{N} \, \mid \, \textit{temp-spec} \\ 
\textit{temp-spec} ::= & \mathbb{N}\{var\} \\
\textit{dims}  ::= & \;\mathbb{N}_{>0} \; \texttt{,} \; \textit{dims} \,
                      \mid \, \mathbb{N}_{>0} \, \mid \, 
                     \texttt{t}
\end{align*}
\caption{Syntax of specifications}
\end{figure}

\begin{figure}
\begin{align*}
(\texttt{a} : \; \texttt{bwd} \; \texttt{depth=1\{b\}} \;
  \texttt{dim=t})
\, & \equiv \,
(\texttt{b} : \; \texttt{bwd} \; \texttt{depth=1\{a\}} \;
  \texttt{dim=t})
\\ 
(\texttt{a} : \; \texttt{fwd} \; \texttt{depth=1\{b\}} \;
  \texttt{dim=t})
\, & \equiv \,
(\texttt{b} : \; \texttt{fwd} \; \texttt{depth=1\{a\}} \;
  \texttt{dim=t})
\\
(a : \texttt{read-once} \; S) \; & \equiv \; (a : \texttt{read-once} \;\,
  \texttt{read-once} \; \, S)
\end{align*}
\caption{Equations on specifications}
\end{figure}

\subsection{Time specifications}

@@

\subsection{Safety and sub-specifications}

\begin{figure}[t]
{\framebox{
\begin{minipage}{0.88\linewidth}
\begin{align*}
\hspace{-0.7em}
\begin{array}{c}
\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
  T}(\textsc{trans}) \\[1.5em]
\dfrac{}{S <: \; \texttt{unspecified dims=}ds}(\top) 
\\[1.5em]
\dfrac{}{\texttt{read-once} \, S <: S}(\textsc{rep})
\\[1.5em]
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} ds \subseteq es \; \wedge \; n \leq m \hspace{3em}}
{\begin{array}{rl}
\texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{fwd} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{bwd} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
  \texttt{depth=}m \; \texttt{dim=}es 
\end{array}}{\footnotesize{(\textsc{cover})}}\hspace{-0.1em}
\end{array}
\end{align*}
\end{minipage}}}
\caption{Definition of sub-specification relation}
\end{figure}

\subsection{Common patterns}

@@

\section{Evaluation}

\section{Discussion}

@@

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}

@@

\newcommand{\bn}{\mathbb{N}}

@@

\newtheorem{lemma}[block]{Lemma}
\newtheorem{proposition}[block]{Proposition}
%\newtheorem{definition}[block]{Definition}

\theoremstyle{definition}

\newtheorem{theorem}[block]{Theorem} 
\newtheorem{remark}[block]{Remark}
\newtheorem{example}[block]{Example}
\newtheorem{definition}[block]{Definition}

@@

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce a specification language for
stencil computations that captures many common forms
of data access pattern, both spatial and temporal (Section~\ref{sec:lang});
\item we detail inference and checking
algorithms for stencil specifications (Section~\ref{sec:analysis});
\item we evaluate our implementation of the approach
in the CamFort tool for Fortran verification, studying
a number of example programs to assess the usefulness
of this approach.
\dnote{insert results here}
\end{itemize}
%

@@

Our specification system is based on the observation
that most forms of array access in numberical code have 
a fixed, statically-determined access pattern. For example, the 
``\emph{five-point stencil}'' on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid 
out-of-bounds access at the edges). We revisit this hypothesis 
in Section~\ref{sec:evaluation} where the inference of
such regular stencil patterns on a corpus of numerical programs (both
small and large). We found that indeed \dnote{..}.

We outline the specification language here. Section~\ref{sec:syntax}
oultines the syntax. Section~\ref{sec:semantics} defines its semantics
via a simple multi-set interpretation over indices. Section~\ref{sec:eqs} provides an
equational theory for specifications via relation ($\equiv$) and a
theory of approximation via relation ($<:$). These equations are then
proven sound with respecto the multi-set semantics of Section~\ref{sec:semantics}.

\subsection{Syntax}
\label{sec:syntax}

Figure~\ref{fig:syntax}
gives the syntax, which we introduces in stages below. 
%%

\begin{figure}
\begin{align*}
\textit{specs} ::= & \; \textit{spec} \; \texttt{;} \; \textit{specs} \, \mid \, \textit{spec} \\
%
\textit{spec} :: = & \; \texttt{v} : S \\[0.5em]
S ::= &  \; \texttt{forward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{backward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{centered} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
& \; \mid \texttt{unspecified} \; [\texttt{dims=}\textit{dims}] \\
& \; \mid \texttt{read-once} \; S \\[0.5em]
\textit{depth} ::= & \; \mathbb{N} \, \mid \, \textit{temp-spec} \\ 
\textit{temp-spec} ::= & \; \mathbb{N}\{var\} \\
\textit{dims} ::= & \; \mathbb{D} \; \texttt{,} \; \textit{dims} \, \mid
                    \, \mathbb{D} \\
\mathbb{D}  ::= & \; \mathbb{N}_{>0} \, \mid \, \texttt{t}
\end{align*}
\caption{Syntax of specifications}
\label{fig:syntax}
\end{figure}

\paragraph{Spatial specifications}

\paragraph{Time specifications}
\dnote{The below is old so probably out of date}

@@

\subsection{A semantic model for specifications}
\label{sec:semantics}

We give a model of our specifications which explains
the range of indices which correspond to the specifications.
The model is based on multi-sets since array indices my be repeated
in an expression.

Figure~\ref{fig:model} describes the model in terms of 
an interpretation function $\interp{-}$ which maps specifications
to multisets of relative indices.

\begin{definition}Relative indices are drawn
from $(\mathbb{Z}_{\bot})^{\mathbb{D}}$, essentially an infinite
tuple of offsets. For example, $(i, j+1)$ corresponds to
the relative index $(0, 1, \bot, \bot, \ldots)$
\end{definition}

The main interpretation function is overloaded on lists of 
variable-spec pairs $\interp{\overline{v : S}}$, 
returning multisets of variable-relative-index pairs. This provides
the top-level definition of the model, with $\interp{-}$ overloaded
on $S$, $\interp{S}$ mqpping to multisets of
relative indices not associated to an array avariable.

\begin{figure}
\begin{align*}
\hspace{-2em}
\begin{array}{c}
\hspace{-1em}
\setlength{\arraycolsep}{0.1em}
\begin{array}{rl}
\interp{\textit{spec}; \textit{specs}} & = \interp{spec} \cup
  \interp{\textit{specs}} \\
%% TEMPORAL SPECS
% \interp{\texttt{v} : \texttt{forward} \; \texttt{depth}=n\texttt{\{u\}} \;
%  \texttt{dims=t}} & =  \{ (\texttt{u}, \textit{ix})^m \mid \forall m,
%                     ix : \mathbb{Z}^{\mathbb{D}}, \textit{ix} \,
%                     \texttt{t} = 0 \} \\[0.4em]
%% SPATIAL SPECS
\interp{\texttt{v} : S} & = \{(\texttt{v}, ix)^m \mid ix^m \in
  \interp{S}\} \\
\end{array}\\[1em]
%\framebox{$\interp{S}$} \\[1em]
\begin{array}{ll}
& \interp{\texttt{reflexive}}
  = \\
& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d = 0 \} \\[0.4em]
& \interp{\texttt{forward} \;\; \texttt{depth}=n \;\;
  \texttt{dims}=\textit{ds}}  = \;\; (\texttt{t} \not\in ds) \\
& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
  \, . \, d \in \textit{ds} \Rightarrow 0 \leq (ix \, d) \leq n \} \\[0.4em]
& \interp{\texttt{backward} \;\; \texttt{depth}=n \;\; \texttt{dims}=\textit{ds}}
  = \; \; (\texttt{t} \not\in ds) \\
& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
  \, . \, d \in \textit{ds} \Rightarrow (-n) \leq (ix \, d) \leq 0 \}
  \\[0.4em]
& \interp{\texttt{read-once} \; S} = \{ ix^1 \mid ix^m \in \interp{S}
  \} 
\end{array}
\end{array}
\end{align*}
\textit{$a^m \in A$ means there are $n$ copies of $a$ in
  the multi-set $A$}
\caption{Multi-set model of specifications}
\label{fig:model}
\end{figure}

\subsection{Equational theory and approximations}
\label{sec:eqs}

Figure~\ref{fig:equations}

\begin{figure}
\begin{align*}
\hspace{-1em}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\overline{\texttt{v} : S} \equiv \overline{\texttt{u} : T}$} \\[1em]
\begin{array}{rl}
(\textsc{coalesceF}) \;\; & (\texttt{a} : \texttt{fwd} \;
                       \texttt{dims}=ds \; \texttt{depth}=n; \\
&  \;\, \texttt{a} :  \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n+1)\\
\equiv \; & \texttt{a} :  \texttt{fwd} \; \texttt{depth}=n+1,
         \texttt{dims}=ds \\[0.75em]
(\textsc{coalesceB}) \;\; & (\texttt{a} : \texttt{bwd} \;
                       \texttt{dims}=ds \; \texttt{depth}=n; \\
&  \;\, \texttt{a} :  \texttt{bwd} \; \texttt{dims}=ds \; \texttt{depth}=n+1) \\
\equiv \; & \texttt{a} :  \texttt{bwd} \; \texttt{depth}=n+1,
         \texttt{dims}=ds \\[0.75em]
(\textsc{idem}) \;\; & (\texttt{a} : S; \texttt{a} : S) \equiv (\texttt{a}
                  : S)
\end{array} \\ \\
\framebox{$\texttt{v} : S \equiv \texttt{v'} : S'$} \\[1em]
\begin{array}{ll}
(\texttt{a} : \; \texttt{bwd} \; \texttt{depth=1\{b\}} \;
  \texttt{dim=t})
\, & \equiv \,
(\texttt{b} : \; \texttt{bwd} \; \texttt{depth=1\{a\}} \;
  \texttt{dim=t})
\\ 
(\texttt{a} : \; \texttt{fwd} \; \texttt{depth=1\{b\}}\;
  \texttt{dim=t})
\, & \equiv \,
(\texttt{b} : \; \texttt{fwd} \; \texttt{depth=1\{a\}} \;
  \texttt{dim=t})
\end{array}
\\ \\
\framebox{$S \equiv S'$} \\[0.75em]
\begin{array}{ll}
(a : \texttt{read-once} \; S) \; & \equiv \; (a : \texttt{read-once} \;\,
  \texttt{read-once} \; \, S)
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

\begin{theorem}[Soundness]
\[
\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

\paragraph{Proof} (see Appendix)

\subsubsection{Safety and sub-specifications}

Figure~\ref{fig:inequations}

\begin{figure}[t]
{\framebox{
\begin{minipage}{0.88\linewidth}
\begin{align*}
\hspace{-0.7em}
\begin{array}{c}
\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
  T}(\textsc{trans}) \\[1.5em]
\dfrac{}{S <: \; \texttt{unspecified dims=}ds}(\top) 
\\[1.5em]
\dfrac{}{\texttt{read-once} \, S <: S}(\textsc{rep})
\\[1.5em]
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} ds \subseteq es \; \wedge \; n \leq m \hspace{3em}}
{\begin{array}{rl}
\texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{fwd} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{bwd} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
  \texttt{depth=}m \; \texttt{dim=}es 
%& \;\, ds \subseteq es \wedge n \leq m 
\\
\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
  \texttt{depth=}m \; \texttt{dim=}es 
\end{array}}
%{\footnotesize{(\textsc{cover})}}
%\hspace{-0.1em}
\end{array}
\end{align*}
\end{minipage}}}
\caption{Definition of sub-specification relation}
\label{fig:inequations}
\end{figure}

\begin{theorem}[Soundness]
\[
\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

\paragraph{Proof} (see Appendix)

\section{Examples and common patterns}

@@

\section{Evaluation}
\label{sec:evaluation}

\section{Discussion}
\label{sec:discussion}

@@

\onecolumn
\appendix

\section{Correctness}

\begin{theorem}[Soundness]
\[
\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

\paragraph{Proof}
\begin{itemize}
\item \textsc{coalesceF}
\begin{align*}
\llbracket & \texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n; \,
  \texttt{v} :  \texttt{fwd} \; \texttt{dims}=ds \;
  \texttt{depth}=n+1)\rrbracket \\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \; \cup \; \{ ix^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq n \}\\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ((ix \, d \leq (n+1)) \vee (ix \,
  d \leq n)) \} \\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \\
& = \interp{\texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \;
  \texttt{depth}=n + 1} \qquad \Box
\end{align*}
\end{itemize}
